### 2015-05-30 papers/histogram/thesis/thesis.tex
changed bibliography referencing style, and formatted thesis for the honors college

%%% bibliography, table of contents
\usepackage[sort&compress,numbers]{natbib} % bibliography options
\bibliographystyle{apsrev4-1}
\usepackage[english]{babel} % allow editing bibliography title
\addto\captionsenglish{ % set bibliography title
  \renewcommand{\contentsname}{Table of Contents}
  \renewcommand\refname{\bf References}}
\usepackage{tocloft}
\renewcommand{\cftsecleader}{\cftdotfill{\cftdotsep}}

@@

\renewcommand{\title}{Optimizing Monte Carlo simulation of the
  square-well fluid}
\renewcommand{\author}{Michael A. Perlin}
\newcommand{\institution}{Oregon State University}
\newcommand{\department}{Department of Physics}
\newcommand{\supervisor}{Dr. David Roundy}
\renewcommand{\date}{15 May 2015}

@@

\fancypagestyle{abstract}{
  \rhead{}
  \lhead{}
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{titlepage}

  \newcommand{\HRule}{\rule{\linewidth}{0.5mm}}

  \center

  \textsc{\LARGE \institution}\\[1.5cm]
  \textsc{\Large \department}\\[0.5cm]

  \HRule \\[0.4cm]
  { \huge \bfseries \title}\\[0.4cm]
  \HRule \\[1.5cm]

  \begin{minipage}{0.4\textwidth}
    \begin{flushleft} \large
      \emph{Author:}\\
      \author
    \end{flushleft}
  \end{minipage}
  ~
  \begin{minipage}{0.4\textwidth}
    \begin{flushright} \large
      \emph{Supervisor:} \\
      \supervisor
    \end{flushright}
  \end{minipage}\\[4cm]

  {\large\date}

  \vfill

\end{titlepage}

@@

\setcounter{page}{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
\label{sec:abstract}

\thispagestyle{abstract}

@@

  \item Construct an initial ``typical'' (i.e. non-ordered) fluid
    configuration. \fixme{explain how?}

@@

The Wang-Landau method is the first published histogram method we
discuss\cite{wang_landau, wang_landau_mod, wang_landau_analysis}. This
method has been used to study a wide variety of systems\fixme{which
  systems? cite}, and is the standard method\fixme{cite} to achieve
broad energy sampling via Monte Carlo simulations.

@@

The first of these parameters, $n$, controls how often to check the
flatness condition $C\sp{H}$ and run updates, if necessary. The value
of $n$ is not too important, so long as it is reasonable for the
system at hand. Too small value of $n$ will waste computation time
determining $C\sp{H}$, while too large of a value will waste time
simulating when $C\sp{H}$ has already been satisfied. In general, $n$
should be proportional to the computation time of $C\sp{H}$, which
scales with energy range of the simulated system. We therefore use
$n=L$, where $L$ is, as described in Section \ref{sec:simple_flat},
the number of energy levels for the given square-well fluid. Other
works typically neglect providing any heuristic or formula for $n$,
reporting a system-independent $n=10^3$\cite{wang_landau_mod} or
$10^5$\cite{wang_landau}.

The next parameter, $f_0$, controls the initial amount by which to
adjust the weight of the current energy after each move. As mentioned
in Algorithm \ref{alg:wang_landau}, one should always have $f_0>1$,
but the appropriate value of $f_0$ will generally depend on the system
at hand. Too large a value of $f_0$ will magnify stochastic error in
initialization, while too small a value of $f_0$ will cause
simulations to take an exceedingly long time to converge. Lacking any
heuristics for assigning $f_0$, we use $f_0=e$ (more precisely, $\ln
f_0=1$), a value commonly suggested by literature\cite{wang_landau}.

The factor $u$ appearing in step \ref{alg:wang_landau_update} of
Algorithm \ref{alg:wang_landau} controls the amount by which to adjust
$f$ when the histogram $H\p{E}$ is sufficiently flat. In principle,
one need not take a root of $f$ at each reset: the general idea behind
Algorithm \ref{alg:wang_landau} requires only $f_k<f_{k-1}$ to
converge (assuming that $C\sp{H}$ can be satisfied) and
$\lim_{k\to\infty}f_k=1$ to terminate. Taking the $u$-th root of $f$
at each increment of $k$ is a natural and convenient iterative means
to satisfy these requirements. Previous works have reported using
values such as $u=2$ \cite{wang_landau, wang_landau_mod} and $u=10$
\cite{wang_landau_analysis}; we use the former value.

@@

The success of Wang-Landau thus relies on some degree of leniency in
the flatness condition, which makes tuning $C\sp{H}$ crucial to the
efficient and effective implementation of this method. Some previous
works\cite{wang_landau} suggest that $C\sp{H}$ should be made as
stringent as possible, cautioning only that some simulations might not
satisfy too strict of a flatness condition in any reasonable amount of
time. These works fail to recognize the reliance of Wang-Landau on the
{\it failure} of $C\sp{H}$ to enforce a perfectly flat histogram. A
commonly suggested flatness condition is that the minimum histogram
value in some specified range of energies be at least some fixed
proportion of the mean histogram value, i.e.
$C\sp{H}:\min\sp{H\p{E}}\ge x\bk{H}_E$, and several papers report
using $x=0.95$. We use this condition with the same value of
$x$. Determining the range of energies over which to evaluate
$C\sp{H}$ is the subject of Section \ref{sec:min_energy}.

@@

Unlike the previous histogram methods, the transition matrix Monte
Carlo method does not use weights, and merely computes the density of
states $D\p{E}$ of a system\cite{tmmc}. The density of states can in
turn be used to determine flat histogram weights via
(\ref{eq:flat_weights}). TMMC introduces a new object: the energy
transition matrix, $T$, whose components $T_{ij}$ are the
probabilities that a system will transition {\it to} a state with
energy $E_i$ {\it from} a given state with energy $E_j$,
i.e. $T_{ij}=P\p{E_j\to E_i}$.

@@

While one could imagine many initialization routines to determine the
transition matrix $T$, we use Algorithm \ref{alg:tmmc}, which is a
modified version of a routine provided in \cite{tmmc}. The primary
motivation behind this algorithm is such: as lower energies are always
more difficult to sample than higher energies, transitions to lower
energies should always be accepted, while transitions to higher
energies should pass a probabilistic acceptance test. To understand
the probability in step \ref{alg:tmmc_prob} of Algorithm
\ref{alg:tmmc} of accepting a move $s_i\to s_f$ for which $E_f>E_i$,
first consider the case of $c=0$ and $T_{\t{min}}\to0$, in which case
the probability becomes
\begin{align}
  P_m\p{s_i\to s_f}=\f{\p{\tilde T_d\p{E_i,\Delta E}}\big/
    \p{\sum_{\Delta E'}\tilde T_d\p{E_i,\Delta E'}}} {\p{\tilde
      T_d\p{E_f,-\Delta E}}\big/ \p{\sum_{\Delta E'}\tilde
      T_d\p{E_f,\Delta E'}}} =\f{T_d\p{E_i,\Delta
      E}}{T_d\p{E_f,-\Delta E}} =\f{P\p{E_i\to E_f}}{P\p{E_f\to E_i}},
  \label{eq:tmmc_prob_simplified}
\end{align}
where $\Delta E=E_f-E_i$ and $P\p{E_m\to E_n}$ is the unbiased
probability of a transition from an energy $E_m$ to an energy
$E_n$. The result in (\ref{eq:tmmc_prob_simplified}) is the
probability of accepting moves $s_i\to s_f$ required for simulation
via Algorithm \ref{alg:tmmc} to yield a flat energy
histogram\cite{tmmc}. Unlike the previous histogram methods, we desire
a flat histogram during initialization via Algorithm \ref{alg:tmmc} in
order to broadly sample the transition histogram $\tilde
T_d\p{E,\Delta E}$, rather than to construct weights $w\p{E}$
directly.

Aside from a minimum temperature $T_{\t{min}}$ and an end condition,
Algorithm \ref{alg:tmmc} contains one free parameter, $c$, used in
step \ref{alg:tmmc_prob} to modify (\ref{eq:tmmc_prob_simplified}).
This parameter, which is our own addition to the algorithm provided in
\cite{tmmc}, makes the rejection of transitions more conservative when
there are too few statistics in the histogram $\tilde T_d$ to be
confident in the current estimate of the transition matrix. Without
this parameter, the TMMC initialization routine can get stuck at
energies which, due to poor collection of statistics in $\tilde T_d$,
appear to be more important than they actually are. In order to make
the effects of $c$ negligible as $\tilde T_d$ collects more statistics
throughout initialization, $c$ should be of order unity, though its
optimal value is generally system-dependent. In our implementation of
Algorithm \ref{alg:tmmc}, we use $c=16$.

@@

All histogram methods introduced in this paper thus far have focused
on determining a weight array $w\p{E}$ for a flat energy histogram
$H\p{E}$, in order to sample and collect statistics on all energies
equally. This motivation considers the quantity of statistics, but not
their quality. Metropolis-Hastings Monte Carlo simulations collect
statistics on low energy states by first getting into these states,
and then rejecting moves which move the system into higher energy
states. Statistics on low energy states will therefore generally be
highly correlated, as they are based on many samples of only a few low
energy regions of a system's energy landscape. The optimized
ensemble\cite{optimized_ensemble} attempts to address the
autocorrelation between low energy samples by finding weights to
maximize the rate at which a simulation makes round trips between low
and high energy states. Maximizing the round trip rate, in turn,
maximizes the rate at which a simulation makes independent,
uncorrelated samples of low energies.

@@

The optimized ensemble method minimizes the integral in
(\ref{eq:walker_current_integral}), thereby maximizing the down-going
walker current $\abs{j}$, by varying the walker density $n$ with the
constraint that $n$ must remain normalized. With the additional
assumption that $\alpha\p{E}$ does not strongly depend on the weights
$w\p{E}$, the authors of \cite{optimized_ensemble} find that the
optimal walker density $n$ is
\begin{align}
  n_{\t{opt}}\propto\f1{\sqrt{\alpha}},
  \label{eq:optimal_walker_density}
\end{align}
where the proportionality is determined by enforcing that $n$ is
normalized.

@@

The algorithm provided in \cite{optimized_ensemble} to construct
weights given by (\ref{eq:optimized_weights}) finds the diffusivity
$\alpha$ by solving for it in
(\ref{eq:walker_current_definition}). This procedure accepts flat
histogram weights, or equivalently the density of states, as an input,
and is not meant to construct weights from scratch. We therefore will
not be comparing the optimized ensemble to other methods in this
paper. We do, however, credit \cite{optimized_ensemble} for motivating
the hybrid DOTMMC method in Section \ref{sec:oetmmc}.

\subsection{The diffusion-optimized transition matrix Monte Carlo
  (DOTMMC) method}
\label{sec:oetmmc}

As well as the density of states, the transition matrix $T\p{E_i,E_j}$
determines the local diffusivity $\alpha\p{E}$ for any given weights
$w\p{E}$. Denoting by $P_b\p{E_i\to E_j}$ the actual, biased
probability of transitioning from $E_i$ to $E_j$ in a single Monte
Carlo move, we can say that
\begin{align}
  P_b\p{E_i\to E_f}=P\p{E_i\to E_f}P_m\p{E_i\to E_f}
  =T_{ji}\max\set{\f{w\p{E_f}}{w\p{E_i}},1},
\end{align}
where $P\p{E_i\to E_j}$ is the probability of attempting the move
$E_i\to E_j$ and $P_m\p{E_i\to E_j}$ is the probability of accepting
that move, should it occur. In terms of a transition histogram $\tilde
T_d\p{E,\Delta E}$ with $E=E_i$ and $\Delta E=E_f-E_i$,
\begin{align}
  P_b\p{E\to E+\Delta E} =\f{\tilde T_d\p{E,\Delta E}w\p{E+\Delta
      E}/w\p{E}} {\sum_{\Delta E'}\tilde T_d\p{E,\Delta
      E'}w\p{E+\Delta E'}/w\p{E}}.
  \label{eq:actual_transition_prob}
\end{align}
We now define
\begin{align}
  \bk{\Delta E}_E=\sum_{\Delta E}\Delta E P_b\p{E\to E+\Delta E}
  =\f{\sum_{\Delta E}\Delta E~\tilde T_d\p{E,\Delta E}
    \max\set{w\p{E+\Delta E}/w\p{E},1}} {\sum_{\Delta E'}\tilde
    T_d\p{E,\Delta E'} \max\set{w\p{E+\Delta E'}/w\p{E},1}}
  \label{eq:<de>}
\end{align}
and
\begin{align}
  \bk{\Delta E^2}_E=\sum_{\Delta E}\Delta E^2 P_b\p{E\to E+\Delta E}
  =\f{\sum_{\Delta E}\Delta E~\tilde T_d\p{E,\Delta E}
    \max\set{w\p{E+\Delta E}/w\p{E},1}} {\sum_{\Delta E'}\tilde
    T_d\p{E,\Delta E'} \max\set{w\p{E+\Delta E'}/w\p{E},1}},
  \label{eq:<de^2>}
\end{align}
in terms of which the diffusivity $\alpha\p{E}$ is
\begin{align}
  \alpha\p{E}=\bk{\Delta E^2}_E -\bk{\Delta E}_E^2.
  \label{eq:microscopic_diffusitivy}
\end{align}
The diffusivity $\alpha\p{E}$ is thus the variance of the change in
energy $\Delta E$ from as single move out of $E$. Reconciling this
definition with that we provide in
(\ref{eq:walker_current_definition}) is considered outside the scope
of this paper. \fixme{say something more convincing? cite?}

The diffusion-optimized transition matrix Monte Carlo (DOTMMC) method,
a hybrid of TMMC and the optimized ensemble, thus first runs Algorithm
\ref{alg:tmmc} to collect a transition histogram and compute flat
histogram weights, then computes the diffusivity $\alpha\p{E}$ via
(\ref{eq:<de>})-(\ref{eq:microscopic_diffusitivy}), and finally
computes optimized weights as given by
(\ref{eq:optimized_weights}). Unlike the optimized ensemble, DOTMMC
does not require any extra simulation routines, and is therefore
straightforward to implement and compare to other histogram methods.

@@

Once we identify an energy $E_-$ satisfying (\ref{eq:min_E_condition})
in an initialization routine for any histogram method, we set weights
$w\p{E<E_-}=e^{-\p{E-E_-}/kT_{\t{min}}}$ with a temperature
$kT_{\t{min}}=0.2\epsilon$. These weights prevent the initialization
routine from wasting time on energies at which we will override the
weight array after initialization anyways. We chose this temperature
because it is comfortably below the temperature at which the
square-well fluid condenses from a gas into a liquid (see Section
\ref{sec:properties}). \fixme{what about pressure?}

@@

The energy histogram in Figure \ref{fig:arrays} shows that TMMC and
DOTMMC spend considerably less time at important low energies than
the simple flat and vanilla Wang-Landau methods. At face value, this
behavior of TMMC and DOTMMC is undesirable, as these are the most
interesting energies for low temperatures, and the most difficult
energies to sample.

High energy histogram counts at low energies mean that a simulation
has collected many statistics on these energies, but the energy
histogram contains no information about how correlated these
statistics are. To determine how effective different simulations are
at collecting uncorrelated statistics, we can look at the pessimistic
sample rate $r_p\p{E}$, which is the mean number of simulation
iterations required to sample a given energy $E$. This rate is roughly
equal to the number of iterations $I$ in a simulation and the
pessimistic sample rate $s_p\p{E}$ by $r_p\p{E}=I/s_p\p{E}$; the
greater the number of iterations $I$, the better this
approximation. Figure \ref{fig:sample_rates} shows pessimistic sample
rates averaged over 30 simulations with $N=25$. Lower pessimistic
sample rates mean that a simulation takes less time to collect
statistics which are guaranteed to be uncorrelated. By this metric,
TMMC and DOTMMC again perform worse than simple flat and vanilla
Wang-Landau.

@@

As with the arrays, the error comparison figures demonstrate the
inability of Wang-Landau to sample a system without a predefined
energy range. The simple flat method, vanilla Wang Landau, and TMMC
seem to give comparable errors, except for some anomalous cases with
$N=18$ and $N=21$ in which simple flat has large errors. We do not
know the reason for these anomalies, but they could occur if the
method terminates before finding energies near $E_-$, in which case it
would get stuck in simulation at low energies with canonical weights
for the minimum temperature $kT_{\t{min}}=0.2\epsilon$. Finally, these
error figures further confirm earlier suspicions of DOTMMC's
underperformance relative to TMMC, despite the fact that it is
supposed to provide an improvement over the TMMC weight array.

@@

I wish to thank, first and foremost my advisor, David Roundy, for
mentoring me and providing me a wealth of knowledge and expertise in
various subjects of physics, numerical methods, and programming
nuances. I would also like thank Paho Lurie-Gregg, with whom I
frequently exchange and discuss codes and ideas, and whose hard
polyhedra Monte Carlo simulations I used as the template for my own
square-well codes. Lastly, I would like to thank thank Eva Morgun, who
prompted me to {\it act} on my textbook enthusiasm for physics, and
get involved in the fun and interesting world of scientific
research. \fixme{Thank Jansen and McIntyre}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\nocite{apsrev41Control} \bibliography{thesis}

>>>>

%%% bibliography, table of contents
\usepackage[sort&compress,numbers]{natbib} % bibliography options
\bibliographystyle{apsrev4-1}
\usepackage[english]{babel} % allow editing bibliography title
\addto\captionsenglish{ % set bibliography title
  \renewcommand{\contentsname}{Table of Contents}
  \renewcommand\refname{References}}
\usepackage{tocloft}
\renewcommand{\cftsecleader}{\cftdotfill{\cftdotsep}}

@@

\usepackage{xspace}

\renewcommand{\title}{Optimizing Monte Carlo simulation of the
  square-well fluid\xspace}
\renewcommand{\author}{Michael A. Perlin\xspace}
\newcommand{\degree}{Honors Baccalaureate of Science
  in Physics\xspace}
\renewcommand{\date}{02 June 2015\xspace}
\newcommand{\email}{\href{mailto:mika.perlin@gmail.com}{mika.perlin@gmail.com}\xspace}

@@

\renewcommand{\titlepage}{
  { \centering

    \title \\
    ~\\~\\
    by \\
    \author \\
    ~\\~\\~\\~\\~\\~\\~\\~\\
    A PROJECT \\
    ~\\~\\
    submitted to \\
    ~\\
    Oregon State University \\
    ~\\
    University Honors College \\
    ~\\~\\~\\~\\~\\~\\
    in partial fulfillment of \\
    the requirements for the \\
    degree of \\
    ~\\~\\
    \degree \\
    (Honors Scholar) \\
    ~\\~\\~\\~\\~\\
    Presented \date \\
    Commencement June 2015 \\
  }
  \newpage
}

@@

\setlength\parindent{0mm}
\pagestyle{empty}

\titlepage

~\newpage

\begin{center}
  AN ABSTRACT OF THE THESIS OF
\end{center}

\author for the degree of \underline{\degree} presented on
\date. Title: \underline{\title}.

~\\~\\~\\~\\

Abstract approved:

~\\
\hrule
\begin{center}
  David Roundy
\end{center}

~\\

~

~\\

Key words: Monte Carlo methods, Square-well fluid, histogram methods

~

Corresponding e-mail address: \email

~\vspace{3cm}

\begin{center}
  \textcopyright Copyright by \author

  \date

  All Rights Reserved
\end{center}

\titlepage

\underline{\degree} project of \author presented on \date.

\vspace{2cm}

APPROVED:

\vspace{2cm}

\hrule ~\\
David Roundy, Mentor, representing the Department of Physics

\vspace{1.5cm}

\hrule ~\\
Henri J. F. Jansen, Committee Member, representing the Department of
Physics

\vspace{1.5cm}

\hrule ~\\
David H. McIntyre, Committee Member, representing the Department of
Physics

\vspace{2.5cm}

\hrule ~\\
Toni Doolen, Dean, University Honors College

\vspace{3cm}

I understand that my project will become part of the permanent
collection of Oregon State University, University Honors College.  My
signature below authorizes release of my project to any reader upon
request.

\vspace{2cm}

\hrule
\begin{center}
  \author, Author
\end{center}

@@

\pagestyle{fancy}
\setlength\parindent{5mm}
\setcounter{page}{1}

@@

  \item Construct an initial ``typical'' (i.e. non-ordered) fluid
    configuration.

@@

The Wang-Landau method is the first published histogram method we
discuss\cite{wang2001determining, belardinelli2007fast,
  zhou2005understanding, landau2004new}. This method has been used to
study a wide variety of systems, and is the standard method to compute
the density of states of a system via Monte Carlo
simulations\cite{landau2004new, yamaguchi2001three,
  parsons2006globule, parsons2006off}.

@@

The first of these parameters, $n$, controls how often to check the
flatness condition $C\sp{H}$ and run updates, if necessary. The value
of $n$ is not too important, so long as it is reasonable for the
system at hand. Too small value of $n$ will waste computation time
determining $C\sp{H}$, while too large of a value will waste time
simulating when $C\sp{H}$ has already been satisfied. In general, $n$
should be proportional to the computation time of $C\sp{H}$, which
scales with energy range of the simulated system. We therefore use
$n=L$, where $L$ is, as described in Section \ref{sec:simple_flat},
the number of energy levels for the given square-well fluid. Other
works typically neglect providing any heuristic or formula for $n$,
reporting a system-independent $n=10^3$\cite{belardinelli2007fast} or
$10^5$\cite{wang2001determining}.

The next parameter, $f_0$, controls the initial amount by which to
adjust the weight of the current energy after each move. As mentioned
in Algorithm \ref{alg:wang_landau}, one should always have $f_0>1$,
but the appropriate value of $f_0$ will generally depend on the system
at hand. Too large a value of $f_0$ will magnify stochastic error in
initialization, while too small a value of $f_0$ will cause
simulations to take an exceedingly long time to converge. Lacking any
heuristics for assigning $f_0$, we use $f_0=e$ (more precisely, $\ln
f_0=1$), a value commonly suggested by
literature\cite{wang2001determining}.

The factor $u$ appearing in step \ref{alg:wang_landau_update} of
Algorithm \ref{alg:wang_landau} controls the amount by which to adjust
$f$ when the histogram $H\p{E}$ is sufficiently flat. In principle,
one need not take a root of $f$ at each reset: the general idea behind
Algorithm \ref{alg:wang_landau} requires only $f_k<f_{k-1}$ to
converge (assuming that $C\sp{H}$ can be satisfied) and
$\lim_{k\to\infty}f_k=1$ to terminate. Taking the $u$-th root of $f$
at each increment of $k$ is a natural and convenient iterative means
to satisfy these requirements. Previous works have reported using
values such as $u=2$ \cite{wang2001determining, belardinelli2007fast}
and $u=10$ \cite{zhou2005understanding}; we use the former value.

@@

The success of Wang-Landau thus relies on some degree of leniency in
the flatness condition, which makes tuning $C\sp{H}$ crucial to the
efficient and effective implementation of this method. Some previous
works\cite{wang2001determining} suggest that $C\sp{H}$ should be made
as stringent as possible, cautioning only that some simulations might
not satisfy too strict of a flatness condition in any reasonable
amount of time. These works fail to recognize the reliance of
Wang-Landau on the {\it failure} of $C\sp{H}$ to enforce a perfectly
flat histogram. A commonly suggested flatness condition is that the
minimum histogram value in some specified range of energies be at
least some fixed proportion of the mean histogram value, i.e.
$C\sp{H}:\min\sp{H\p{E}}\ge x\bk{H}_E$, and several papers report
using $x=0.95$. We use this condition with the same value of
$x$. Determining the range of energies over which to evaluate
$C\sp{H}$ is the subject of Section \ref{sec:min_energy}.

@@

Unlike the previous histogram methods, the transition matrix Monte
Carlo method does not use weights, and merely computes the density of
states $D\p{E}$ of a system\cite{wang2002transition}. The density of
states can in turn be used to determine flat histogram weights via
(\ref{eq:flat_weights}). TMMC introduces a new object: the energy
transition matrix, $T$, whose components $T_{ij}$ are the
probabilities that a system will transition {\it to} a state with
energy $E_i$ {\it from} a given state with energy $E_j$,
i.e. $T_{ij}=P\p{E_j\to E_i}$.

@@

While one could imagine many initialization routines to determine the
transition matrix $T$, we use Algorithm \ref{alg:tmmc}, which is a
modified version of a routine provided in
\cite{wang2002transition}. The primary motivation behind this
algorithm is such: as lower energies are always more difficult to
sample than higher energies, transitions to lower energies should
always be accepted, while transitions to higher energies should pass a
probabilistic acceptance test. To understand the probability in step
\ref{alg:tmmc_prob} of Algorithm \ref{alg:tmmc} of accepting a move
$s_i\to s_f$ for which $E_f>E_i$, first consider the case of $c=0$ and
$T_{\t{min}}\to0$, in which case the probability becomes
\begin{align}
  P_m\p{s_i\to s_f}=\f{\p{\tilde T_d\p{E_i,\Delta E}}\big/
    \p{\sum_{\Delta E'}\tilde T_d\p{E_i,\Delta E'}}} {\p{\tilde
      T_d\p{E_f,-\Delta E}}\big/ \p{\sum_{\Delta E'}\tilde
      T_d\p{E_f,\Delta E'}}} =\f{T_d\p{E_i,\Delta
      E}}{T_d\p{E_f,-\Delta E}} =\f{P\p{E_i\to E_f}}{P\p{E_f\to E_i}},
  \label{eq:tmmc_prob_simplified}
\end{align}
where $\Delta E=E_f-E_i$ and $P\p{E_m\to E_n}$ is the unbiased
probability of a transition from an energy $E_m$ to an energy
$E_n$. The result in (\ref{eq:tmmc_prob_simplified}) is the
probability of accepting moves $s_i\to s_f$ required for simulation
via Algorithm \ref{alg:tmmc} to yield a flat energy
histogram\cite{wang2002transition}. Unlike the previous histogram
methods, we desire a flat histogram during initialization via
Algorithm \ref{alg:tmmc} in order to broadly sample the transition
histogram $\tilde T_d\p{E,\Delta E}$, rather than to construct weights
$w\p{E}$ directly.

Aside from a minimum temperature $T_{\t{min}}$ and an end condition,
Algorithm \ref{alg:tmmc} contains one free parameter, $c$, used in
step \ref{alg:tmmc_prob} to modify (\ref{eq:tmmc_prob_simplified}).
This parameter, which is our own addition to the algorithm provided in
\cite{wang2002transition}, makes the rejection of transitions more
conservative when there are too few statistics in the histogram
$\tilde T_d$ to be confident in the current estimate of the transition
matrix. Without this parameter, the TMMC initialization routine can
get stuck at energies which, due to poor collection of statistics in
$\tilde T_d$, appear to be more important than they actually are. In
order to make the effects of $c$ negligible as $\tilde T_d$ collects
more statistics throughout initialization, $c$ should be of order
unity, though its optimal value is generally system-dependent. In our
implementation of Algorithm \ref{alg:tmmc}, we use $c=16$.

@@

All histogram methods introduced in this paper thus far have focused
on determining a weight array $w\p{E}$ for a flat energy histogram
$H\p{E}$, in order to sample and collect statistics on all energies
equally. This motivation considers the quantity of statistics, but not
their quality. Metropolis-Hastings Monte Carlo simulations collect
statistics on low energy states by first getting into these states,
and then rejecting moves which move the system into higher energy
states. Statistics on low energy states will therefore generally be
highly correlated, as they are based on many samples of only a few low
energy regions of a system's energy landscape. The optimized
ensemble\cite{trebst2004optimizing} attempts to address the
autocorrelation between low energy samples by finding weights to
maximize the rate at which a simulation makes round trips between low
and high energy states. Maximizing the round trip rate, in turn,
maximizes the rate at which a simulation makes independent,
uncorrelated samples of low energies.

@@

The optimized ensemble method minimizes the integral in
(\ref{eq:walker_current_integral}), thereby maximizing the down-going
walker current $\abs{j}$, by varying the walker density $n$ with the
constraint that $n$ must remain normalized. With the additional
assumption that $\alpha\p{E}$ does not strongly depend on the weights
$w\p{E}$, the authors of \cite{trebst2004optimizing} find that the
optimal walker density $n$ is
\begin{align}
  n_{\t{opt}}\propto\f1{\sqrt{\alpha}},
  \label{eq:optimal_walker_density}
\end{align}
where the proportionality is determined by enforcing that $n$ is
normalized.

@@

The algorithm provided in \cite{trebst2004optimizing} to construct
weights given by (\ref{eq:optimized_weights}) finds the diffusivity
$\alpha$ by solving for it in
(\ref{eq:walker_current_definition}). This procedure accepts flat
histogram weights, or equivalently the density of states, as an input,
and is not meant to construct weights from scratch. We therefore will
not be comparing the optimized ensemble to other methods in this
paper. We do, however, credit \cite{trebst2004optimizing} for
motivating the hybrid OETMMC method in Section \ref{sec:oetmmc}.

\subsection{The hybrid OETMMC method}
\label{sec:oetmmc}

As well as the density of states, the transition matrix $T\p{E_i,E_j}$
determines the local diffusivity $\alpha\p{E}$ for any given weights
$w\p{E}$. Denoting by $P_b\p{E_i\to E_j}$ the actual, biased
probability of transitioning from $E_i$ to $E_j$ in a single Monte
Carlo move, we can say that
\begin{align}
  P_b\p{E_i\to E_f}=P\p{E_i\to E_f}P_m\p{E_i\to E_f}
  =T_{ji}\max\set{\f{w\p{E_f}}{w\p{E_i}},1},
\end{align}
where $P\p{E_i\to E_j}$ is the probability of attempting the move
$E_i\to E_j$ and $P_m\p{E_i\to E_j}$ is the probability of accepting
that move, should it occur. In terms of a transition histogram $\tilde
T_d\p{E,\Delta E}$ with $E=E_i$ and $\Delta E=E_f-E_i$,
\begin{align}
  P_b\p{E\to E+\Delta E} =\f{\tilde T_d\p{E,\Delta E}w\p{E+\Delta
      E}/w\p{E}} {\sum_{\Delta E'}\tilde T_d\p{E,\Delta
      E'}w\p{E+\Delta E'}/w\p{E}}.
  \label{eq:actual_transition_prob}
\end{align}
We now define
\begin{align}
  \bk{\Delta E}_E=\sum_{\Delta E}\Delta E P_b\p{E\to E+\Delta E}
  =\f{\sum_{\Delta E}\Delta E~\tilde T_d\p{E,\Delta E}
    \max\set{w\p{E+\Delta E}/w\p{E},1}} {\sum_{\Delta E'}\tilde
    T_d\p{E,\Delta E'} \max\set{w\p{E+\Delta E'}/w\p{E},1}}
  \label{eq:<de>}
\end{align}
and
\begin{align}
  \bk{\Delta E^2}_E=\sum_{\Delta E}\Delta E^2 P_b\p{E\to E+\Delta E}
  =\f{\sum_{\Delta E}\Delta E~\tilde T_d\p{E,\Delta E}
    \max\set{w\p{E+\Delta E}/w\p{E},1}} {\sum_{\Delta E'}\tilde
    T_d\p{E,\Delta E'} \max\set{w\p{E+\Delta E'}/w\p{E},1}},
  \label{eq:<de^2>}
\end{align}
in terms of which the diffusivity $\alpha\p{E}$ is
\begin{align}
  \alpha\p{E}=\bk{\Delta E^2}_E -\bk{\Delta E}_E^2.
  \label{eq:microscopic_diffusitivy}
\end{align}
The diffusivity $\alpha\p{E}$ is thus the variance of the change in
energy $\Delta E$ from as single move out of $E$. Reconciling this
definition with that we provide in
(\ref{eq:walker_current_definition}) is considered outside the scope
of this paper.

The OETMMC method, a hybrid of TMMC and the optimized ensemble, first
runs Algorithm \ref{alg:tmmc} to collect a transition histogram and
compute flat histogram weights, then computes the diffusivity
$\alpha\p{E}$ via (\ref{eq:<de>})-(\ref{eq:microscopic_diffusitivy}),
and finally computes optimized weights as given by
(\ref{eq:optimized_weights}). Unlike the optimized ensemble, OETMMC
does not require any extra simulation routines, and is therefore
straightforward to implement and compare to other histogram methods.

@@

Once we identify an energy $E_-$ satisfying (\ref{eq:min_E_condition})
in an initialization routine for any histogram method, we set weights
$w\p{E<E_-}=e^{-\p{E-E_-}/kT_{\t{min}}}$ with a temperature
$kT_{\t{min}}=0.2\epsilon$. These weights prevent the initialization
routine from wasting time on energies at which we will override the
weight array after initialization anyways. We chose this temperature
because it is comfortably below the temperature at which the
square-well fluid condenses from a gas into a liquid in simulation
(see Section \ref{sec:properties}).

@@

The energy histogram in Figure \ref{fig:arrays} shows that TMMC and
OETMMC spend considerably less time at important low energies than the
simple flat and vanilla Wang-Landau methods. At face value, this
behavior of TMMC and OETMMC is undesirable, as these are the most
interesting energies for low temperatures, and the most difficult
energies to sample.

High energy histogram counts at low energies mean that a simulation
has collected many statistics on these energies, but the energy
histogram contains no information about how correlated these
statistics are. To determine how effective different simulations are
at collecting uncorrelated statistics, we can look at the pessimistic
sample rate $r_p\p{E}$, which is the mean number of simulation
iterations required to sample a given energy $E$. This rate is roughly
equal to the number of iterations $I$ in a simulation and the
pessimistic sample rate $s_p\p{E}$ by $r_p\p{E}=I/s_p\p{E}$; the
greater the number of iterations $I$, the better this
approximation. Figure \ref{fig:sample_rates} shows pessimistic sample
rates averaged over 30 simulations with $N=25$. Lower pessimistic
sample rates mean that a simulation takes less time to collect
statistics which are guaranteed to be uncorrelated. By this metric,
TMMC and OETMMC again perform worse than simple flat and vanilla
Wang-Landau.

@@

As with the arrays, the error comparison figures demonstrate the
inability of Wang-Landau to sample a system without a predefined
energy range. The simple flat method, vanilla Wang Landau, and TMMC
seem to give comparable errors, except for some anomalous cases with
$N=18$ and $N=21$ in which simple flat has large errors. We do not
know the reason for these anomalies, but they could occur if the
method terminates before finding energies near $E_-$, in which case it
would get stuck in simulation at low energies with canonical weights
for the minimum temperature $kT_{\t{min}}=0.2\epsilon$. Finally, these
error figures further confirm earlier suspicions of OETMMC's
underperformance relative to TMMC, despite the fact that it is
supposed to provide an improvement over the TMMC weight array.

@@

I wish to thank, first and foremost my advisor, David Roundy, for
mentoring me and providing me a wealth of knowledge and expertise in
various subjects of physics, numerical methods, and programming
nuances. I would also like thank Paho Lurie-Gregg, with whom I
frequently exchange and discuss codes and ideas, and whose hard
polyhedra Monte Carlo simulations I used as the template for my own
square-well codes. Lastly, I would like to thank thank Eva Morgun, who
prompted me to {\it act} on my textbook enthusiasm for physics, and
get involved in the fun and interesting world of scientific
research.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\nocite{apsrev41Control} \bibliography{thesis} \lhead{}
